Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota

### 2010-02-25 Paper.tex
Messed with titles.

\section{Overview of Occupancy Modeling}

    \subsection{What is Occupancy Rate?}
    Occupancy has two different definitions. It can mean ``the probability that
    a randomly selected site or sampling unit in an area of interest is occupied
    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
    proportion of area, patches, or sample units that is occupied''
    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
    population size, of a species. The number of members of a species is not
    important for either definition of occupancy.

@@

    \subsection{Different Ways of Sampling Data}
    In order to determine occupancy rate, the data is often collected in the
    form of a presence-absence survey. This means that observers visit sites and
    look for the animals themselves, tracks, vocalizations, or some other
    indicator that the animal is present. In some cases, camera traps and sound
    recording are used instead of human observers \cite{MacKenzie2006}.

@@

    \subsection{Types of Occupancy Models}
    All of the models discussed in this paper are single-species, single-season
    models. It is possible to modify all of these models to be used for multiple
    seasons or multiple species, but it would make them more complicated.

@@

    \subsection{Markov Chains and Testing the Independence Assumption}
    \label{shat}
    First we considered using Markov Chains to come up with a long term
    probability of detection, which we called \(r\). We let \(p\) be the number
    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
    \(1\) as a fraction of all successive pairs that start with a \(1\). These
    probabilities can be represented by the matrix:
    \[M=\begin{bmatrix}
        p & 1-p \\
        1-q & q
    \end{bmatrix}\text{.}\]

@@

\section{Theoretical Stuff}

    \subsection{Bayesian Statistics}
    Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\

@@

\section{Standard Hierarchical Model}

    \subsection{Our Hierarchical Model}
    For our model, we specified uniform \((0, 1)\) prior distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked and therefore are estimated. The covariates of this
    model are: observer, date, and the number of days since the last snow event.

    \subsection{Simulated Data}
    We simulated both balanced and unbalanced datasets meant for the standard
    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
    observer, and \(E\) for each observer.  Balanced data sets, in this context,
    contain the same number of flights by the same observers after each snow
    event.  Naturally, unbalanced data sets do not.

@@

    \subsection{Obtaining Convergence on the Actual Data}
    Recall that our standard hierarchical model contains the following
    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability of
    laying down a track given occupancy), \(p\) (the probability of detecting a
    track given that it is present, which is unique to each observer and makes
    \(1-p\) the false negative rate of detecting a track), and \(E\) (the false
    positive rate of detecting a track, which is unique to each observer).

@@

    \subsection{Future Work}
    Additional observations are more helpful than additional snow events at
    first, although both are much more helpful than the addition of an observer.
    Once the numbers of snow events and observations after each snow event both
    surpass three, they improve the model's estimates at roughly equal rates.
    Note, however, that increasing the number of snow events is likely to be far
    less challenging than increasing the number of observations after each snow
    event at that point, especially in southeastern Minnesota.

>>>>

\section{Overview of occupancy modeling}

    \subsection{What is occupancy rate?}
    Occupancy has two different definitions. It can mean ``the probability that
    a randomly selected site or sampling unit in an area of interest is occupied
    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
    proportion of area, patches, or sample units that is occupied''
    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
    population size, of a species. The number of members of a species is not
    important for either definition of occupancy.

@@

    \subsection{Different ways of sampling data}
    In order to determine occupancy rate, the data is often collected in the
    form of a presence-absence survey. This means that observers visit sites and
    look for the animals themselves, tracks, vocalizations, or some other
    indicator that the animal is present. In some cases, camera traps and sound
    recording are used instead of human observers \cite{MacKenzie2006}.

@@

    \subsection{Types of occupancy models}
    All of the models discussed in this paper are single-species, single-season
    models. It is possible to modify all of these models to be used for multiple
    seasons or multiple species, but it would make them more complicated.

@@

    \subsection{Testing the independence assumption}
    \label{shat}
    First we considered using Markov Chains to come up with a long term
    probability of detection, which we called \(r\). We let \(p\) be the number
    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
    \(1\) as a fraction of all successive pairs that start with a \(1\). These
    probabilities can be represented by the matrix:
    \[M=\begin{bmatrix}
        p & 1-p \\
        1-q & q
    \end{bmatrix}\text{.}\]

@@

\section{Bayesian modeling theory and practice}

    \subsection{Bayesian statistics}
    Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\

@@

\section{Standard hierarchical model}

    \subsection{Our hierarchical model}
    For our model, we specified uniform \((0, 1)\) prior distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked and therefore are estimated. The covariates of this
    model are: observer, date, and the number of days since the last snow event.

    \subsection{Simulated data}
    We simulated both balanced and unbalanced datasets meant for the standard
    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
    observer, and \(E\) for each observer.  Balanced data sets, in this context,
    contain the same number of flights by the same observers after each snow
    event.  Naturally, unbalanced data sets do not.

@@

    \subsection{Obtaining convergence on the data}
    Recall that our standard hierarchical model contains the following
    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability of
    laying down a track given occupancy), \(p\) (the probability of detecting a
    track given that it is present, which is unique to each observer and makes
    \(1-p\) the false negative rate of detecting a track), and \(E\) (the false
    positive rate of detecting a track, which is unique to each observer).

@@

    \subsection{Future work}
    Additional observations are more helpful than additional snow events at
    first, although both are much more helpful than the addition of an observer.
    Once the numbers of snow events and observations after each snow event both
    surpass three, they improve the model's estimates at roughly equal rates.
    Note, however, that increasing the number of snow events is likely to be far
    less challenging than increasing the number of observations after each snow
    event at that point, especially in southeastern Minnesota.
